---
title: "Assessing Stem Cell Therapeutics in Murine Models"
author: "Giles Carlos, Nathan Gin, Alexander Nathaneal, Vinh Nguyen, Owen Sitiabudi"
date: "`r Sys.Date()`"
output:
  pdf_document: 
    number_sections: true
  html_document:
    number_sections: true
---

```{r echo = FALSE, include = FALSE}
library(tidyverse)
library(kableExtra)
library(knitr)
library(dplyr)
library(readr)
library(ggplot2)

post_injury <- read_csv(here::here("data/post_injury.csv"))
```


# Introduction and Problem Statement



# Related Work

DeepLabCut 

# Data Sets

```{r echo = FALSE}
kable(head(post_injury, 10),
      escape = T, caption = "Example of First 10 Rows of Manually Input Data Set Obtained from Ladder Beam Task",
      align = 'l') %>% 
  kable_styling(latex_options = "HOLD_position") %>% 
  row_spec(0, bold = TRUE)
```

^[Mathis, A., Mamidanna, P., Cury, K.M. et al. DeepLabCut: markerless pose estimation of user-defined body parts with deep learning. Nat Neurosci 21, 1281–1289 (2018). https://doi.org/10.1038/s41593-018-0209-y]

^[Weber, R.Z., Mulders, G., Kaiser, J. et al. Deep learning-based behavioral profiling of rodent stroke recovery. BMC Biol 20, 232 (2022). https://doi.org/10.1186/s12915-022-01434-9]

^[Aljovic, A., Zhao, S., Chahin, M. et al. A deep learning-based toolbox for Automated Limb Motion Analysis (ALMA) in murine models of neurological disorders. Commun Biol 5, 131 (2022). https://doi.org/10.1038/s42003-022-03077-6]




# Overall Technical Approach

# Software

# Experiments and Evaluation

# Notebook Description

# Members Participation

```{r echo = FALSE}
covariates <- c("task 1")

giles <- c("100%")
nathan <- c("100%")
alex <- c("100%")
vinh <- c("100%")
owen <- c("100%")

workload_summary <- data.frame(covariates,
                               linebreak(giles),
                               linebreak(nathan), 
                               linebreak(alex),
                               linebreak(vinh),
                               linebreak(owen))
                               
colnames(workload_summary) <- c("Task", "Giles", "Nathan", "Alex", "Vinh", "Owen")

kable(workload_summary, 
      escape = T, 
      caption = "Percentage of workload across group members",
      align = 'l') %>% 
  kable_styling(latex_options = "HOLD_position") %>% 
  row_spec(0, bold = TRUE)
```


# Discussion and Conclusion

*What did you learn about the methods and algorithms you worked with? What did you learn about their strengths? And their limitations?*

> test text

*What ended up being harder than you expected in your project? What was surprising about your project?*

> test

*What other lessons did you learn, expected or unexpected (e.g., perhaps about the tools you used, if you used anything out of the ordinary?*

> test text

*If you were in charge of a research lab, what ideas and directions might you invest in over the next year or two to try to make major progress on this problem? Feel free to be speculative in discussing possible future directions.*

> test text
 
